% paper.tex — Sutra paper LaTeX source with anonymization macros.
%
% Two builds from the same source via the \ifanon switch:
%   pdflatex \def\anontrue{} \input{paper.tex}    -> anonymized (NeurIPS double-blind)
%   pdflatex \input{paper.tex}                    -> full version (with author + repo URL)
%
% The CI/CD workflow (.github/workflows/paper-pdf.yml — TODO) will
% drive both builds and publish them to the docs site.
%
% The substantive paper content lives in paper/paper.md (Markdown
% for clawRxiv submission). This LaTeX file is the NeurIPS-shaped
% skeleton; the body is currently a thin wrapper that includes the
% Markdown content via pandoc-converted intermediate (paper.tex.body
% generated at build time). Full NeurIPS-template integration (the
% specific year's neurips_*.sty file) is deferred until the deadline
% is announced and the template URL is confirmed.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{xcolor}
\usepackage{geometry}
\geometry{margin=1in}

% ====================================================================
% Anonymization switch — set \anontrue at build time for the NeurIPS
% double-blind PDF. Default (no flag) builds the named version.
% ====================================================================
\newif\ifanon
% Default: not anonymized. Override at command line:
%   pdflatex "\def\anontrue{} \input{paper.tex}"
% (or via the build workflow's -DANON=1 conditional logic)

\ifanon
  % Anonymized version — NeurIPS double-blind safe.
  \title{From Learned Displacements to Learned Matrices:\\
    A Programming Language for Vector-Symbolic Computation\\
    in Frozen Embedding Spaces}
  \author{Anonymous Authors}
  \newcommand{\repo}{[anonymized repository]}
  \newcommand{\siblingcartography}{[anonymized companion repository]}
  \newcommand{\sutradburl}{[anonymized companion repository]}
  \newcommand{\authoremail}{[anonymized]}
\else
  % Full version — for clawRxiv preprint and post-acceptance camera-ready.
  \title{From Learned Displacements to Learned Matrices:\\
    Sutra, a Programming Language for Vector-Symbolic Computation\\
    in Frozen Embedding Spaces}
  \author{Emma Leonhart\\
    \texttt{immanuelleleonhart@gmail.com}}
  \newcommand{\repo}{\url{https://github.com/EmmaLeonhart/Sutra}}
  \newcommand{\siblingcartography}{\url{https://github.com/EmmaLeonhart/latent-space-cartography}}
  \newcommand{\sutradburl}{\url{https://github.com/EmmaLeonhart/SutraDB}}
  \newcommand{\authoremail}{\texttt{immanuelleleonhart@gmail.com}}
\fi

\date{2026-04-30 draft}

\begin{document}
\maketitle

\begin{abstract}
Frozen general-purpose language-model embedding spaces encode
relational structure as vector arithmetic — a property
established empirically in the cartography line of work that this
paper follows from. We argue this empirical foundation suggests a
three-step research arc: (1) isolate regular displacements in the
embedding space; (2) consolidate them into canonical operational
primitives; (3) generalize from rank-zero displacements
(translations) to learned-matrix role operators. Step 1 is prior
published work; this paper presents step 2 plus the design and
implementation of \ifanon{}a programming language\else{}\textbf{Sutra}\fi,
a typed, purely functional programming language whose compile target
is a single tensor-op graph over the frozen embedding substrate.
\ifanon{}The system\else{}Sutra\fi{} is a working compiler today:
parser, type checker, codegen, runtime; three demonstration programs
plus loop demonstrations execute end-to-end with all expected outputs
correct. The language design positions step 3 (learned-matrix
binding) as an explicit extension point with rotation binding as the
working substrate today. The paper's contribution is the language
that operationalizes the empirical findings, plus an honest account
of which substrate-purity claims are shipped and which remain
pending fixes.
\end{abstract}

\section{Note on this LaTeX file}

The substantive paper content is maintained in \texttt{paper.md}
(Markdown, suitable for clawRxiv and pandoc-to-anything). This
\texttt{paper.tex} is the NeurIPS-shaped skeleton: it sets the
title/author/repo URLs via the \texttt{\textbackslash{}ifanon}
switch and stages the abstract for review-ready PDF generation.
The body sections will be either:

\begin{itemize}
  \item \textbf{Hand-converted to LaTeX} when the NeurIPS deadline
    is announced and the year-specific template is integrated.
    Anchors: see \texttt{paper.md} sections 1--9 for content.
  \item \textbf{Generated by pandoc} at build time
    (\texttt{pandoc paper.md -t latex -o paper.tex.body}), with
    this file becoming a thin wrapper. Easier but loses fine
    typographic control.
\end{itemize}

The CI/CD workflow at \texttt{.github/workflows/paper-pdf.yml}
(TODO, sub-item 1d) will drive both anonymized and named PDF
builds and publish them as workflow artifacts.

\section{Repository pointers}

\begin{itemize}
  \item Sutra repository: \repo{}
  \item Cartography (step 1, prior work): \siblingcartography{}
  \item SutraDB (sibling triplestore): \sutradburl{}
  \item Reproducibility checklist: \texttt{paper/REPRODUCE.md}
  \item Skill description: \texttt{paper/SKILL.md}
  \item Project devlog (1407 commits, 2026-03-13 to 2026-04-30):
    \texttt{DEVLOG.md}
\end{itemize}

\section*{Why this is a skeleton, not a finished LaTeX paper}

For \texttt{paper.md} on clawRxiv, the Markdown source is what
ships and is complete. For NeurIPS, a year-specific LaTeX template
is required and the year's deadline+template aren't confirmed yet.
Rather than guess the template (NeurIPS evolves it annually),
this skeleton stages the anon-macro infrastructure so the future
template-integration step is a short edit, not a full rewrite.

\end{document}
